Colorimetric Detection of pH Strips

### 2017-12-10 CS189/paper/paper.tex
fixed paper typos

\begin{abstract}
In this paper, we apply several machine learning techniques and models for the purpose of classifying and estimating the pH of solutions given raw image data of pH test strips. We replicate and critique much of the work performed by Mutlu et. al [1], who used LS-SVM and claimed to achieve $100\%$ classification. We believe that this high accuracy was the result of duplicating physical pH strips samples between training and validation datasets; this is problematic because the noise between different strips of the same class was not accounted for, and samples were essentially duplicated as pre-processing mitigated orientation variability efforts. We further find that regression is a more suitable approach for this domain, as pH values are on a continuous, logarithmic scale, and decimal differences in value can have significant biological consequences. In this spirit, we find that mean squared regression errors as low as $\sim 0.033$ are achievable.
\end{abstract}

@@

\section{Introduction}
\IEEEPARstart{G}{rowing} cell cultures for industrial-scale chemical output is a very difficult task. Large bioreactors cost tens of thousands of dollars and require complex, specialized machinery to operate. An alternative idea is to make use of many 5-gallon buckets, with a film on top to allows air to pass through the system, but blocks liquids. With a pump inside the bucket to mix nutrients with the cells, we essentially have a low-cost bioreactor. We need to test the pH of each bucket periodically to determine when to feed the cells. This can be done in a cost-effective manner by using paper-based strips that are briefly submerged into the solution. This method of bioreactors can only be profitable if there are many buckets - on the order of hundreds or thousands. Therefore, we need an efficient, automated method of determining pH from the test strips. The literature on this suggest that in practice, smartphones are used as the source of these pictures.

@@

Using a kNN regression model on the unscaled data across the four experiments, mean squared errors of $0.034$, $0.109$, $0.292$, and $0.949$ were achieved for $k=2$, $2$, $2$, $3$, respectively (Fig. 8). The lowest error on this model nearly as reliable as the lowest error achieved using ridge regression ($0.033$). After scaling and using PCA on the kNN regression model, the mean squared error for both the third and fourth experiments was greatly lowered (Fig. 9). For $k=11$, $12$ these models achieved mean squared errors of $0.197$ and $0.392$, a significant improvement over the other regression models. Concerning the first two experiments, which did not employ color rotation, kNN achieved mean squared errors of $0.046$ and $0.074$ with $k=3$, $3$ neighbors respectively.

\section{Discussion}
\subsection{Classification}
The best classification model for the unmodified color extracted data, as in our Experiment 1, was LDA with parameters… This result matched the $100\%$ accuracy that Mutlu’s best model, their LS-SVM, achieved. To a great extent, an LDA model is more conceptually sensible than an LS-SVM model for this application. This is because each pH value class has a single ground truth strip that corresponds to it--all photos taken are simply noisy samples from this ground truth strip. Therefore, it makes sense that the LDA model performs so well--it estimates the mean features of the ground truth strip and estimates a covariance matrix surrounding each mean. In other words, the mechanism of the model directly corresponds with the way data is generated. On the other hand, LS-SVM does not offer the same resemblance and in fact, the presence of outliers in a training set could significantly skew the decision boundary produced since they will incur a sizeable loss and force the hyperplane to move.

@@

In our experiments with discriminative classification models, we chose to use one-hot encodings as is standard for working with multiple classes; one-hot encoding ensures that there is no ordering among the classes. However, for our application, the pH value classes actually do have a natural and logical ordering (a pH value of 7 is closer to a value of 6 than 2, for example). This would be an avenue of future exploration, and we hypothesize that not using one-hot encoding may further increase our accuracy since nearby classes only differ by one or two color panels and thus, even in the presence of noisy lighting conditions, the model will have a better chance of recognizing the variations between neighboring classes as opposed to those between each pair of classes. Further, this would prevent the model from ever being in a situation in which it must decide between two classes at opposite ends of the spectrum--with pH as a log-based scale, an incredibly inaccurate prediction like this could have disastrous implications and it would be best to avoid such a possibility.

>>>>

\begin{abstract}
In this paper, we apply several machine learning techniques and models for the purpose of classifying and estimating the pH of solutions given raw image data of pH test strips. We replicate and critique much of the work performed by Mutlu et. al [1], who used LS-SVM and claimed to achieve 100\% classification. We believe that this high accuracy was the result of duplicating physical pH strips samples between training and validation datasets; this is problematic because the noise between different strips of the same class was not accounted for, and samples were essentially duplicated as pre-processing mitigated orientation variability efforts. We further find that regression is a more suitable approach for this domain, as pH values are on a continuous, logarithmic scale, and decimal differences in value can have significant biological consequences. In this spirit, we find that mean squared regression errors as low as $\sim$0.033 are achievable.
\end{abstract}

@@

\section{Introduction}
\IEEEPARstart{G}{rowing} cell cultures for industrial-scale chemical output is a very difficult task. Large bioreactors cost tens of thousands of dollars and require complex, specialized machinery to operate. An alternative idea is to make use of many 5-gallon buckets, with a film on top that allows air to pass through the system, but blocks liquids. With a pump inside the bucket to mix nutrients with the cells, we essentially have a low-cost bioreactor. We need to test the pH of each bucket periodically to determine when to feed the cells. This can be done in a cost-effective manner by using paper-based strips that are briefly submerged into the solution. This method of bioreactors can only be profitable if there are many buckets - on the order of hundreds or thousands. Therefore, we need an efficient, automated method of determining pH from the test strips. The literature on this suggest that in practice, smartphones are used as the source of these pictures.

@@

Using a kNN regression model on the unscaled data across the four experiments, mean squared errors of $0.034$, $0.109$, $0.292$, and $0.949$ were achieved for $k=2$, $2$, $2$, $3$, respectively (Fig. \ref{Rknn}). The lowest error on this model nearly as reliable as the lowest error achieved using ridge regression ($0.033$). After scaling and using PCA on the kNN regression model, the mean squared error for both the third and fourth experiments was greatly lowered (Fig. \ref{Rknn_pca}). For $k=11$, $12$ these models achieved mean squared errors of $0.197$ and $0.392$, a significant improvement over the other regression models. Concerning the first two experiments, which did not employ color rotation, kNN achieved mean squared errors of $0.046$ and $0.074$ with $k=3$, $3$ neighbors respectively.

\section{Discussion}
\subsection{Classification}
The best classification model for the unmodified color extracted data, as in our Experiment 1, was LDA with PCA pre-processing using 12 principal components. This result matched the $100\%$ accuracy that Mutlu \textquotesingle s best model, their LS-SVM, achieved. To a great extent, an LDA model is more conceptually sensible than an LS-SVM model for this application. This is because each pH value class has a single ground truth strip that corresponds to it -- all photos taken are simply noisy samples from this ground truth strip. Therefore, it makes sense that the LDA model performs so well -- it estimates the mean features of the ground truth strip and estimates a covariance matrix surrounding each mean. In other words, the mechanism of the model directly corresponds with the way data is generated. On the other hand, LS-SVM does not offer the same resemblance and in fact, the presence of outliers in a training set could significantly skew the decision boundary produced since they will incur a sizeable loss and force the hyperplane to move.

@@

In our experiments with discriminative classification models, we chose to use one-hot encodings as is standard for working with multiple classes; one-hot encoding ensures that there is no ordering among the classes. However, for our application, the pH value classes actually do have a natural and logical ordering (a pH value of 7 is closer to a value of 6 than 2, for example). This would be an avenue of future exploration, and we hypothesize that not using one-hot encoding may further increase our accuracy since nearby classes only differ by one or two color panels and thus, even in the presence of noisy lighting conditions, the model will have a better chance of recognizing the variations between neighboring classes as opposed to those between each pair of classes. Further, this would prevent the model from ever being in a situation in which it must decide between two classes at opposite ends of the spectrum -- with pH as a log-based scale, an incredibly inaccurate prediction like this could have disastrous implications and it would be best to avoid such a possibility.
